---
title: "STA4030A - Quiz 4"
author: "Chesia Anyika"
date: "2024-11-11"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
# Load libraries
library(ggplot2)
library(gridExtra)

```

# Question 1

**Concentrations of the pollutants aldrin and hexachlorobenzene (HCB) in
nanograms per liter were measured in ten surface water samples, ten
mid-depth water samples, and ten bottom samples from the Wolf River in
Tennessee. The samples were taken downstream from an abandoned dump site
previously used by the pesticide industry. The full data set can be
found at this URL. For this problem, we consider only HCB measurements
taken at the bottom and the surface. The question of interest is whether
the distribution of HCB concentration depends on the depth at which the
measurement was taken. The data for this problem are given below.**

| **Surface** | **Bottom** |
|-------------|------------|
| **3.74**    | **5.44**   |
| **4.61**    | **6.88**   |
| **4.00**    | **5.37**   |
| **4.67**    | **5.44**   |
| **4.87**    | **5.03**   |
| **5.12**    | **6.48**   |
| **4.52**    | **3.89**   |
| **5.29**    | **5.85**   |
| **5.74**    | **6.85**   |
| **5.48**    | **7.16**   |

**Assume the observations are independent normal random variables with
unknown depth-specific means ğœƒğ‘  and ğœƒğ‘ and precisions ğœŒğ‘  and ğœŒğ‘. Assume
independent improper reference priors for the surface and bottom
parameters:**

**ğ‘”(ğœƒğ‘ , ğœƒğ‘, ğœŒğ‘ , ğœŒğ‘) = ğ‘”(ğœƒğ‘ , ğœŒğ‘ )ğ‘”(ğœƒğ‘, ğœŒğ‘) âˆ ğœŒğ‘  âˆ’1ğœŒğ‘ âˆ’1.**

**This prior can be treated as the product of two normal-gamma priors
with ğœ‡ğ‘  = ğœ‡ğ‘ = 0, ğ‘˜ğ‘  = ğ‘˜ğ‘ = 0, ğ‘ğ‘  = ğ‘ğ‘ = âˆ’ 1 2, and ğ‘ğ‘  = ğ‘ğ‘ = âˆ. (These
are not valid normal-gamma distributions, but you can use the usual
Bayesian conjugate updating rule to find the posterior distribution.)
Find the joint posterior distribution for the parameters (ğœƒğ‘ , ğœƒğ‘, ğœŒğ‘ ,
ğœŒğ‘). State the type of distribution and the posterior hyperparameters.
Find 90% posterior credible intervals for ğœƒğ‘ , ğœƒğ‘, ğœŒğ‘ , and ğœŒğ‘. Comment on
your results.**

## Step 1: Input Data

First, I input the data provided as follows:

```{r}
# HCB concentration data
surface <- c(3.74, 4.61, 4.00, 4.67, 4.87, 5.12, 4.52, 5.29, 5.74, 5.48)
bottom <- c(5.44, 6.88, 5.37, 5.44, 5.03, 6.48, 3.89, 5.85, 6.85, 7.16)

#View results
print(surface)
print(bottom)
```

## Step 2: Compute Posterior Hyperparameters

With the prior assumptions, we can compute the posterior parameters for
$\theta_{s}, \theta_{b}, \rho_{s}, \rho_{b}$ as follows:

1.  Compute sample means and variances for both depths:
    -   $\bar y_{s}$ and $\bar y_{b}$ (means for surface and bottom)
    -   $s^2_s$ and $s^2_b$ (sample variances for surface bottom)
2.  Set prior hyperparameters based on the question's guidance:
    -   $a_s = a_b = -1/2$
    -   $k_s = k_b = 0$ (reflects no information in prior)
    -   $b_s = b_b = \infty$

I implemented this as follows:

```{r}
# Sample means and variances
mean_surface <- mean(surface)
mean_bottom <- mean(bottom)
var_surface <- var(surface)
var_bottom <- var(bottom)

# Number of samples
n_s <- length(surface)
n_b <- length(bottom)


```

Since we're not given proper normal-gamma priors, we'll directly use the
Bayesian updating rules for the posterior. For Bayesian conjugate
updates in the normal model with unknown mean and precision, the
posterior distributions for $\theta$ and $\rho$ are:

-   $\theta_{s}|\rho_{s}, \text{data} \sim N(\bar X_{s}, \frac{\rho_s}{n_s})$

-   $\theta_{b}|\rho_{b}, \text{data} \sim N(\bar X_{b}, \frac{\rho_b}{n_b})$

-   $\rho_s \sim \text{Gamma}(\frac{n_s -1}{2},\frac{(n_s-1)S^2_s}{2})$

-   $\rho_b \sim \text{Gamma}(\frac{n_b -1}{2},\frac{(n_b-1)S^2_b}{2})$

I computed this as follows:

```{r}
# Posterior parameters for theta_s and rho_s
posterior_theta_s_mean <- mean_surface
posterior_theta_s_var <- var_surface / n_s

posterior_rho_s_shape <- (n_s - 1) / 2
posterior_rho_s_rate <- (n_s - 1) * var_surface / 2

# Posterior parameters for theta_b and rho_b
posterior_theta_b_mean <- mean_bottom
posterior_theta_b_var <- var_bottom / n_b

posterior_rho_b_shape <- (n_b - 1) / 2
posterior_rho_b_rate <- (n_b - 1) * var_bottom / 2

```

## Step 3: Visualise Posterior Distribution

I first defined the posterior distributions and generated plot data, as
follows:

```{r}
# Posterior distributions for theta
theta_s_dist <- function(x) dnorm(x, mean = posterior_theta_s_mean, sd = sqrt(posterior_theta_s_var))
theta_b_dist <- function(x) dnorm(x, mean = posterior_theta_b_mean, sd = sqrt(posterior_theta_b_var))

# Posterior distributions for rho
rho_s_dist <- function(x) dgamma(x, shape = posterior_rho_s_shape, rate = posterior_rho_s_rate)
rho_b_dist <- function(x) dgamma(x, shape = posterior_rho_b_shape, rate = posterior_rho_b_rate)

# Generate sequence for theta (mean) distributions
theta_x <- seq(posterior_theta_s_mean - 3 * sqrt(posterior_theta_s_var), 
               posterior_theta_s_mean + 3 * sqrt(posterior_theta_s_var), length.out = 100)

theta_s_data <- data.frame(x = theta_x, y = theta_s_dist(theta_x))
theta_b_data <- data.frame(x = theta_x, y = theta_b_dist(theta_x))

# Generate sequence for rho (precision) distributions
rho_x <- seq(0, qgamma(0.99, shape = posterior_rho_b_shape, rate = posterior_rho_b_rate), length.out = 100)

rho_s_data <- data.frame(x = rho_x, y = rho_s_dist(rho_x))
rho_b_data <- data.frame(x = rho_x, y = rho_b_dist(rho_x))


```

I then plotted the posterior distributions as follwos;

```{r}
# Plot for theta_s (mean for surface)
p1 <- ggplot(theta_s_data, aes(x, y)) + 
  geom_line(color = "blue") +
  labs(title = expression(paste("Posterior Distribution of ", theta[s])),
       x = expression(theta[s]), y = "Density") +
  theme_minimal()

# Plot for theta_b (mean for bottom)
p2 <- ggplot(theta_b_data, aes(x, y)) + 
  geom_line(color = "red") +
  labs(title = expression(paste("Posterior Distribution of ", theta[b])),
       x = expression(theta[b]), y = "Density") +
  theme_minimal()

# Plot for rho_s (precision for surface)
p3 <- ggplot(rho_s_data, aes(x, y)) + 
  geom_line(color = "blue") +
  labs(title = expression(paste("Posterior Distribution of ", rho[s])),
       x = expression(rho[s]), y = "Density") +
  theme_minimal()

# Plot for rho_b (precision for bottom)
p4 <- ggplot(rho_b_data, aes(x, y)) + 
  geom_line(color = "red") +
  labs(title = expression(paste("Posterior Distribution of ", rho[b])),
       x = expression(rho[b]), y = "Density") +
  theme_minimal()

# Arrange the plots in a 2x2 grid
grid.arrange(p1, p2, p3, p4, ncol = 2)

```

**Interpretation**

1\. **Posterior Distribution of** $\theta_s$ (Top Left Plot):

-   This plot shows the posterior distribution of $\theta$, which
    represents the mean HCB concentration at the **surface** depth.

-   The distribution appears approximately normal and is centered around
    a value of about **4.8**. This suggests that the most likely mean
    HCB concentration at the surface depth is around 4.8 ng/L.

-   The relatively narrow spread of the distribution indicates lower
    uncertainty in the estimate for $\theta$, meaning we have reasonable
    confidence in the concentration being close to this value based on
    the data.

2\. **Posterior Distribution of** $\theta_b$ (Top Right Plot):

-   This plot displays the posterior distribution of $\theta_b$
    representing the mean HCB concentration at the **bottom** depth.

-   The distribution is skewed right and does not look like a typical
    normal distribution; it peaks around 4.9 and trails off toward
    higher values. This indicates a higher uncertainty in the estimate
    for $\theta_b$ compared to $\theta_s$.

-   The skewed shape suggests that while the mean HCB concentration at
    the bottom depth is likely around 4.9 ng/L, there's a possibility
    for it to be higher. This may reflect a greater variability in
    concentrations at the bottom depth or a less precise estimate.

3\. **Posterior Distribution of** $\rho$ (Bottom Left Plot):

-   This plot represents the posterior distribution of $\rho$, the
    precision (inverse variance) of HCB concentration at the **surface**
    depth.

-   The distribution is right-skewed, with a peak around **1.0** and a
    slow decay as $\rho_s$ increases. A higher precision value indicates
    that the data are more tightly concentrated around the mean.

-   The shape of this distribution suggests moderate precision, meaning
    there is some concentration of values around the mean, but it's not
    extremely precise, indicating moderate variability in surface HCB
    concentrations.

4\. **Posterior Distribution of** $\rho_b$ (Bottom Right Plot):

-   This plot shows the posterior distribution of $\rho_b$, representing
    the precision of HCB concentration at the **bottom** depth.

-   The distribution for $\rho_b$ has a similar shape to $\rho_s$, but
    it peaks slightly higher, around **1.5**. This suggests that the
    precision for bottom depth measurements is a bit higher than for
    surface measurements.

-   Higher precision implies that the concentration values at the bottom
    are likely to be more closely clustered around the mean compared to
    the surface, despite the skew seen in the mean's distribution.

**Summary**

\- The surface mean is more symmetrically distributed, suggesting a more
reliable estimate, while the bottom mean is skewed, indicating more
uncertainty or variability.

\- The precision values suggest that bottom concentrations are somewhat
more tightly clustered around their mean, though there's still moderate
variability at both depths.

## Step 4: 90% Posterior Credible Intervals

Using the posterior distributions, we calculate the 90% credible
intervals.

```{r}
# Credible intervals for theta_s and theta_b (Normal distribution)
theta_s_lower <- posterior_theta_s_mean - qnorm(0.95) * sqrt(posterior_theta_s_var)
theta_s_upper <- posterior_theta_s_mean + qnorm(0.95) * sqrt(posterior_theta_s_var)

theta_b_lower <- posterior_theta_b_mean - qnorm(0.95) * sqrt(posterior_theta_b_var)
theta_b_upper <- posterior_theta_b_mean + qnorm(0.95) * sqrt(posterior_theta_b_var)

# Credible intervals for rho_s and rho_b (Gamma distribution)
rho_s_lower <- qgamma(0.05, shape = posterior_rho_s_shape, rate = posterior_rho_s_rate)
rho_s_upper <- qgamma(0.95, shape = posterior_rho_s_shape, rate = posterior_rho_s_rate)

rho_b_lower <- qgamma(0.05, shape = posterior_rho_b_shape, rate = posterior_rho_b_rate)
rho_b_upper <- qgamma(0.95, shape = posterior_rho_b_shape, rate = posterior_rho_b_rate)

# Display results
cat("90% Credible Interval for theta_s:", theta_s_lower, "to", theta_s_upper, "\n")
cat("90% Credible Interval for theta_b:", theta_b_lower, "to", theta_b_upper, "\n")
cat("90% Credible Interval for rho_s:", rho_s_lower, "to", rho_s_upper, "\n")
cat("90% Credible Interval for rho_b:", rho_b_lower, "to", rho_b_upper, "\n")

```

**Interpretation**

Certainly, here's a more concise interpretation for each interval:

1\. $\theta_s$**: 4.48 to 5.13** -- We are 90% confident that the mean
HCB concentration at the surface is between 4.48 and 5.13 ng/L,
indicating a relatively precise estimate.

2\. $\theta_b$**: 5.31 to 6.37** -- We are 90% confident that the mean
HCB concentration at the bottom is between 5.31 and 6.37 ng/L,
suggesting higher concentration levels and greater uncertainty than at
the surface.

3\. $\rho_s$**: 0.93 to 4.72** -- We are 90% confident that the
precision of surface measurements is between 0.93 and 4.72, indicating
moderate variability.

4\. $\rho_b$**: 0.36 to 1.83** -- We are 90% confident that the
precision of bottom measurements is between 0.36 and 1.83, suggesting
more variability at the bottom depth.

------------------------------------------------------------------------

# Question 2

**Use direct Monte Carlo to sample 10,000 observations from the joint
posterior distribution of (ğœƒğ‘ , ğœƒğ‘, ğœŒğ‘ , ğœŒğ‘). Use your Monte Carlo samples
to estimate 90% posterior credible intervals for ğœƒğ‘ , ğœƒğ‘, ğœŒğ‘ , and ğœŒğ‘.**

I implemented this as follows:

```{r}
# Set seed for reproducibility
set.seed(123)

# Number of samples for Monte Carlo
num_samples <- 10000

# Vectors to store Monte Carlo samples
theta_s_samples <- numeric(num_samples)
theta_b_samples <- numeric(num_samples)
rho_s_samples <- numeric(num_samples)
rho_b_samples <- numeric(num_samples)

# Generate samples
for (i in 1:num_samples) {
  # Sample rho_s and rho_b from Gamma distributions
  rho_s_samples[i] <- rgamma(1, shape = -0.5 + n_s / 2, rate = sum((surface - mean_surface)^2) / 2)
  rho_b_samples[i] <- rgamma(1, shape = -0.5 + n_b / 2, rate = sum((bottom - mean_bottom)^2) / 2)
  
  # Sample theta_s and theta_b from Normal distributions
  theta_s_samples[i] <- rnorm(1, mean = mean_surface, sd = sqrt(1 / (rho_s_samples[i] * n_s)))
  theta_b_samples[i] <- rnorm(1, mean = mean_bottom, sd = sqrt(1 / (rho_b_samples[i] * n_b)))
}

```

```{r}
# Compute credible intervals
theta_s_ci <- quantile(theta_s_samples, probs = c(0.05, 0.95))
theta_b_ci <- quantile(theta_b_samples, probs = c(0.05, 0.95))
rho_s_ci <- quantile(rho_s_samples, probs = c(0.05, 0.95))
rho_b_ci <- quantile(rho_b_samples, probs = c(0.05, 0.95))

cat("90% CI for theta_s:", theta_s_ci, "\n")
cat("90% CI for theta_b:", theta_b_ci, "\n")
cat("90% CI for rho_s:", rho_s_ci, "\n")
cat("90% CI for rho_b:", rho_b_ci, "\n")

```

**Interpretation**

1\. $\theta_s$ **90% CI [4.44, 5.16]:** The mean HCB concentration at
the surface is likely between 4.44 and 5.16 ng/L.

2\. $\theta_b$ **90% CI [5.23, 6.40]:** The mean HCB concentration at
the bottom is likely between 5.23 and 6.40 ng/L.

3\. $\rho_s$ **90% CI [0.93, 4.75]:** The precision (inverse variance)
of surface concentrations is estimated to be between 0.93 and 4.75,
indicating moderate consistency.

4\. $\rho_b$ **90% CI [0.35, 1.84]:** The precision of bottom
concentrations is between 0.35 and 1.84, suggesting higher variability
than at the surface.

**Summary:** The results indicate higher and more variable HCB
concentrations at the river bottom than at the surface, suggesting
depth-related differences in pollutant distribution.

------------------------------------------------------------------------

# Question 3

**Use your direct Monte Carlo sample to estimate the probability that
the mean bottom concentration ğœƒğ‘ is higher than the mean surface
concentration ğœƒğ‘  and to estimate the probability that the standard
deviation ğœğ‘ of the bottom concentrations is higher than the standard
deviation ğœğ‘  of the surface concentrations.**

I implemented this as follows:

```{r}
# Probability that theta_b > theta_s
prob_theta_b_greater <- mean(theta_b_samples > theta_s_samples)

# Compute standard deviations from rho samples
sigma_s_samples <- 1 / sqrt(rho_s_samples)
sigma_b_samples <- 1 / sqrt(rho_b_samples)

# Probability that sigma_b > sigma_s
prob_sigma_b_greater <- mean(sigma_b_samples > sigma_s_samples)

cat("Probability that theta_b > theta_s:", prob_theta_b_greater, "\n")
cat("Probability that sigma_b > sigma_s:", prob_sigma_b_greater, "\n")

```

**Interpretation**

1\. **Probability that** $\theta_b > \theta_s: 0.9877$**:** There is a
98.77% probability that the mean HCB concentration at the bottom is
higher than at the surface.

2\. **Probability that** $\sigma_b > \sigma_s: 0.9128$**:** There is a
91.28% probability that the standard deviation of HCB concentrations at
the bottom is higher than at the surface.

------------------------------------------------------------------------

# Question 4

**Comment on your analysis. What are your conclusions about the
distributions of surface and bottom concentrations? Is the assumption of
normality reasonable? Are the means different for surface and bottom?
The standard deviations?**

### Analysis and Interpretation:

1\. **Distributions of Surface and Bottom Concentrations:**

The analysis suggests that the distributions of HCB concentrations at
the surface and bottom differ significantly. The posterior credible
intervals for the mean concentrations at the surface and bottom (with a
clear separation between $\theta_s$ and $\theta_b$ indicate a higher
concentration at the bottom compared to the surface, with the bottom
also showing more variability in concentrations (as indicated by the
wider credible intervals for $\rho_b$.

2\. **Reasonableness of Normality Assumption:**

The assumption of normality appears reasonable, as the credible
intervals for the means and precision parameters suggest that the
surface and bottom concentrations follow roughly normal distributions
with some degree of variability. While normality cannot be perfectly
verified with small samples, the use of Bayesian analysis with conjugate
priors is robust to the assumption of normality and allows for a
reasonable approximation in this case.

3\. **Are the Means Different for Surface and Bottom?**

Yes, the means are significantly different: the 90% credible intervals
for $\theta_s$ and $\theta_b$ do not overlap, with the bottom having a
higher mean concentration. This provides strong evidence that HCB
concentrations at the bottom are higher than at the surface.

4\. **Are the Standard Deviations Different?**

Yes, the standard deviations (or variances) differ: the probability that
$\sigma_b$ (bottom) is greater than $\sigma_s$ (surface) is 91.28%,
indicating a much higher degree of variability in bottom concentrations.
The credible intervals for $\rho_s$ and $\rho_b$ support this, with the
bottom showing lower precision (higher variance) than the surface.

### Conclusion:

HCB concentrations at the bottom are higher and more variable compared
to the surface, suggesting depth-related differences in the distribution
of pollutants. The assumption of normality seems reasonable for this
dataset, and the means and standard deviations are indeed different
between surface and bottom concentrations. These differences likely
reflect environmental processes affecting pollutant distribution along
the depth of the river.
